---
title: "Oyster Populations in Chesapeake Bay"
author: "tracking_oystRs"
date: "2023-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(ggmap)
library(usmap)
```

# Introduction
Info to include:
Oysters have been a staple of the Chesapeake Bay region
Eastern oyster species
Choptank river location
Significance of oysters = cultural, economic, ecological

```{r regional map, echo=F, message=F,fig.dim=c(4,2),fig.cap="The Chesapeake Bay region"}
plot_usmap(include=c("DE","NJ","MD","VA","PA"))
```

# Methods
```{r ocean temperature, echo=F, message=F}
temp_data <- read_csv("water_temp_data.csv")
```

```{r pH data, echo=F, message=F}
ph_data <- read_csv("ph_data.csv")
```

```{r salinity data, echo=F, message=F}
sal_data <- read_csv("salinity_data.csv")
```

```{r oyster data, echo=F, message=F}
dat <- read_csv("oyster_data.csv")
```

```{r joining the data, echo=F, message=F}

```

```{r plotting joined data, echo=F, message=F}
```
# Results

# Discussion

# Author Contributions
Audrey:
Conor:
Jessica:
Letacianna:
Susanne:

Notes:
calculate temperature anomaly (mean temperature by day)
  Water quality crisis per year from a mean or standard
Add nitrogen or chlorophyll (use interactions for things that correlate)
Response variables: which few you want to predict
  Mortality
  
Visualization - response variable over time
Delegate tasks

Planning:
What factor do we want to look at: young oysters, sexually mature oysters, or market ready oysters
