---
title: "Oyster Populations in Chesapeake Bay"
author: "tracking_oystRs"
date: "2023-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(ggmap)
library(usmap)
library(ggplot2)
library(gridExtra)
```

# Introduction
Info to include:
Oysters have been a staple of the Chesapeake Bay region
Eastern oyster species
Choptank river location
Significance of oysters = cultural, economic, ecological

# The Cultural Significance of Oysters
Oysters have long played a crucial role in the cultural history of the Chesapeake Bay. According to a cultural model assessment conducted by Paolisso and Dery [@paolisso2010cultural], the restoration of oyster populations in the region is a topic of significant importance. The study explores various oyster restoration alternatives and provides insights into the cultural perspectives surrounding these efforts.  The cultural significance of oysters in the Chesapeake Bay extends beyond their ecological impact. Paolisso and Dery [@paolisso2010cultural] discuss how different restoration approaches are perceived within the local communities. Oysters are not merely a commodity; they are deeply embedded in the cultural fabric, shaping traditions and community practices. Understanding the cultural dimensions of oyster restoration is essential for developing sustainable and community-supported initiatives in the Chesapeake Bay region.

```{r regional map, echo=F, message=F,fig.dim=c(4,2),fig.cap="The Chesapeake Bay region"}
plot_usmap(include=c("DE","NJ","MD","VA","PA"))
```

# Methods
```{r ocean temperature, echo=F, message=F}
temp_data <- read_csv("water_temp_data.csv")
```

```{r pH data, echo=F, message=F}
ph_data <- read_csv("ph_data.csv")
```

```{r salinity data, echo=F, message=F}
sal_data <- read_csv("salinity_data.csv")
```

```{r oyster data, echo=F, message=F}
oys_data <- read_csv("oyster_data.csv")
ggplot(oys_data,aes(x=Year,y=oyster_units))+geom_point()
```

```{r joining the data, echo=F, message=F,fig.cap="Average salinity, average pH, average water temperature, and temperature divergence from the average plotted against year"}
environ <- merge(x=sal_data,y=ph_data,by="Year",all.x=TRUE)%>%
  left_join(temp_data,by="Year")

plot1 <- ggplot(environ,aes(x=Year,y=salinity_Average))+geom_point()
plot2 <- ggplot(environ,aes(x=Year,y=ph_Average))+geom_point()
plot3 <- ggplot(environ,aes(x=Year,y=water_temp_Average))+geom_point()
plot4 <- ggplot(environ,aes(x=Year,y=water_temp_Divergence))+geom_point()

grid.arrange(plot1,plot2,plot3,plot4)

dat <- environ %>%
  left_join(oys_data,by="Year")

colnames(dat)[6] <- "oyster_units"
dat <- dat[-c(7)] #removing unnecessary column of repeated data

```

```{r plotting joined data, echo=F, message=F}
ggplot(dat, aes(x=Year))+geom_point(y=dat$oyster_units)
```

Visualize each variable against oyster numbers
Linear mixed modeling to find best model

# Results

# Discussion

# Author Contributions
Audrey:

Conor:

Jessica:

Letacianna:

Susanne:

Notes:
calculate temperature anomaly (mean temperature by day)
  Water quality crisis per year from a mean or standard
Add nitrogen or chlorophyll (use interactions for things that correlate)
Response variables: which few you want to predict
  Mortality
  
Visualization - response variable over time
Delegate tasks

Planning:
What factor do we want to look at: young oysters, sexually mature oysters, or market ready oysters
